Analysis-ready VCF at Biobank scale using Zarr

### 2024-06-07 paper.tex

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets, 
and has been instrumental in projects such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
% Glow~\cite{glow2024} has recently emerged as a similar approach
% to performing distributed analytics on genetic variation data.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
the details of its MatrixTable format are not documented
or intended for external reuse. 
% cite? https://dev.hail.is/t/matrixtable-file-format-reference/173/6
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
awshealthomics2024,microsoftgenomics2024}
or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

@@

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped by
access level within chunks (padding partially filled 
chunks as needed and using an appropriate sample mask). 
Providing client applications direct access to 
the data over HTTP
and delegating access control to the cloud provider
makes custom web APIs~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
largely unnecessary.

@@

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} provide powerful abstractions
for large-scale data analysis and can be
used to perform interactive analysis
of large VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which optimised kernels are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis. These abstractions will extend, hopefully, to
include more convenient primitives for genomics workflows through
projects like sgkit in the future.

>>>>

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets, 
and has been instrumental in projects such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
the details of its MatrixTable format are not documented
or intended for external reuse. 
% cite? https://dev.hail.is/t/matrixtable-file-format-reference/173/6
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
awshealthomics2024,microsoftgenomics2024}
or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

@@

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped by
access level within chunks (padding partially filled 
chunks as needed and using an appropriate sample mask). 
Providing client applications direct access to 
the data over HTTP
and delegating access control to the cloud provider
makes custom web APIs~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
largely unnecessary in this setting.

@@

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working 
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}  
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters. 
This scaling is achieved by distributing calculations over 
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created 
to facilitate work on a scalable genetics toolkit for 
Python~\cite{sgkit2024} built on Xarray. 
While the high-level facilities for distributed 
computation provided by Xarray 
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here 
illustrate that working at the lowest level, 
by sequentially applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers 
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.
